---
title: "GE 585 Activity 2"
author: "John Foster"
output: html_document
---



1. Generate plots of the logistic growth model at r = 1.95, 2.05, 2.5, and 2.8 Describe the trajectory observed in each case.

```{r}
library(ecoforecastR)
r <- c(1.95, 2.05, 2.5, 2.8)
K = 10        ## carrying capacity      
n0 = .1       ## initial population size
NT = 30       ## number of time steps to simulate
time = 1:NT

n = matrix(n0, length(r), NT)    ## vector to store results

for(g in 1:length(r)){
  for(t in 2:NT){
    n[g,t] = n[g,t-1] + r[g]*n[g,t-1]*(1-n[g,t-1]/K)
  }
}

plot(time,n[1,],
     ylim = c(0,max(n)*1.1),
     type = "l",
     bty = 'l',
     cex.lab = 1.5,
     xlab = "Time",
     ylab = "Population Size")
for(i in 2:length(r)){
  lines(time, n[i,],
        col = i,
        bty = 'l')
}
legend("bottomright",
       legend = r,
       col = 1:length(r),
       lwd = 1,
       title = "Rate of increase")
```

All populations reach carrying capacity at roughly the same time, however we see that as `r` increases the magnitude of occilations around `K` increase. We also see that the number of time points between peaks and valleys becomes more variable as `r` increases.


2. Choose another probability distribution and generate graphs of the probability density function, the cumulative distribution function, the quantile function, and a histogram of samples from that distribution.

```{r}
x = 0:50
lambda <- 12
plot(x,dpois(x, lambda),type='l')  		## that’s a lowercase “L” for “line”
abline(v = lambda)					## add a line to indicate the mean (“v” is for “vertical”)
lines(x,dpois(x,lambda+4),col=2)			## try changing the mean (“col” sets the color)
abline(v=lambda+4,col=2)
lines(x,dpois(x,lambda+20),col=3)			## try changing the mean and standard dev
abline(v=lambda+20,col=3)

plot(x,ppois(x, lambda),type='l')
abline(v=lambda)
lines(x,ppois(x, lambda+4),col=2)
abline(v=lambda+4,col=2)
lines(x,ppois(x, lambda+20),col=3)
abline(v=lambda+20,col=3)

p = seq(0,1,by=0.01)
plot(p,qpois(p, lambda),type='l',ylim=range(x))    
abline(h=lambda)					
lines(p,qpois(p, lambda+4),col=2)
abline(h=lambda+4,col=2)
lines(p,qpois(p, lambda+20),col=3)
abline(h=lambda+20,col=3)

n = c(10,100,1000,10000)	# sequence of sample sizes
for(i in 1:4){			# loop over these sample sizes
  hist(rpois(n[i], lambda),main=n[i],probability=TRUE)  
				#here breaks defines number of bins in the histogram
  lines(x,dpois(x, lambda),col=2)
}


```


3. Numerically transform a lognormal(meanlog=0,sdlog=0.5) through sin(x) using Monte Carlo simulation. Include histograms of the original and transformed distributions. Report the mean, median, and 95% CI for both distributions and indicate these values on the histograms.

```{r}
n.mcmc <- 5000
dlnorm <- rlnorm(n.mcmc, 0, 0.5)
y <- sin(dlnorm)

hist(dlnorm, main = "Original lognormal",breaks=40)
abline(v=quantile(dlnorm,c(0.025,0.5,0.975)),lty=c(2,1,2),lwd=3,col="orange")
abline(v=mean(dlnorm),col="red",lwd=3,lty=3)

hist(y,main="Transformed distribution",breaks=40)
abline(v=quantile(y,c(0.025,0.5,0.975)),lty=c(2,1,2),lwd=3,col="orange")
abline(v=mean(y),col="red",lwd=3,lty=3)

```

Lognormal:

Mean: `r mean(dlnorm)`

Median: `r median(dlnorm)`

Lower 95% CI: `r quantile(dlnorm, 0.025)`

Upper 95% CI: `r quantile(dlnorm, 0.975)`

Sin Transformed:

Mean: `r mean(y)`

Median: `r median(y)`

Lower 95% CI: `r quantile(y, 0.025)`

Upper 95% CI: `r quantile(y, 0.975)`


```{r}
r.sd = 0.2     ## standard deviation on r
K.sd = 1.0     ## standard deviation on K
NE = 1000      ## Ensemble size

n = matrix(n0,NE,NT)   # storage for all simulations
rE = rnorm(NE,r,r.sd)  # sample of r
KE = rnorm(NE,K,K.sd)  # sample of K
for(i in 1:NE){        # loop over samples
  for(t in 2:NT){      # for each sample, simulate throught time
    nn = n[i,t-1] + rE[i]*n[i,t-1]*(1-n[i,t-1]/KE[i])
    n[i,t] <- max(0, nn)
  }
}

n.stats = apply(n,2,quantile,c(0.025,0.5,0.975))

ciEnvelope <- function(x,ylo,yhi,col="lightgrey",...){
  polygon(cbind(c(x, rev(x), x[1]), c(ylo, rev(yhi),
  ylo[1])), border = NA,col=col,...) 
}
```

### Problems

4. Plot histograms of the samples of r and K used for the simulation.

```{r}
hist(rE, 
     main = "Samples of r",
     xlab = "r")
hist(KE,
     main = "Samples of K",
     xlab = "K")
```

5. Plot a sample of 10 different trajectories from your ensemble (on one graph).

```{r}
nsamp <- 10
draw <- sample.int(nrow(n), nsamp)

plot(time, n[draw[1],], 
     type = "l",
     ylab = "Population Size",
     xlab = "Time",
     main = "10 Ensemble Trajectories")
for(i in 2:10){
  lines(time, n[draw[i],],
        col = i)
}

```

6. Plot a histogram of your population forecast at time = 15.

```{r}
hist(n[,15])
```

7. Plot the median trajectory and 95% CI.

8. Add a 50% CI (i.e. 25% to 75%) to the plot. Note that you'll have to both compute the summary statistics for this interval and plot the envelope in a different color.

```{r}
stats.50 = apply(n,2,quantile,c(0.25,0.75))
plot(time,n.stats[2,],
     type = "l",
     xlab = "Time",
     ylab = "Population")
ciEnvelope(time,n.stats[1,],n.stats[3,], col = "lightblue")
ciEnvelope(time,stats.50[1,],stats.50[2,], col = "lightgreen")
lines(time,n.stats[2,])
```





Extra Credit: Initial conditions
--------------------------------

The approach for simulating uncertainty in the initial conditions is very similar to the approach used for the parameter uncertainty.  As in Chapter 2, we'll assume that the initial condition is distributed as a lognormal to ensure that we never draw negative values. For this example we'll assume a standard deviation of 0.6 and an intrinsic growth rate of 0.3

```{r}
r = 0.3
n0.sd = 0.6
n0s = rlnorm(NE,log(n0),n0.sd)
n = matrix(n0s,NE,NT)
for(i in 1:NE){
  for(t in 2:NT){
    n[i,t] = n[i,t-1] + r*n[i,t-1]*(1-n[i,t-1]/K)
  }
}
```

### Problems

9. Plot the median & 95% interval.

```{r}
n.stats = apply(n,2,quantile,c(0.025,0.5,0.975))
plot(time,n.stats[2,],
     type = "l",
     xlab = "Time",
     ylab = "Population")
ciEnvelope(time,n.stats[1,],n.stats[3,], col = "lightblue")
lines(time,n.stats[2,])
```

10. Repeat with r equal to 1.95, 2.05, and 2.8

```{r}
r = c(1.95, 2.05, 2.8)
n0.sd = 0.6
n0s = rlnorm(NE,log(n0),n0.sd)
n = array(n0s,dim = c(NE,NT,3))
for(g in 1:3){
  for(i in 1:NE){
    for(t in 2:NT){
      n[i,t,g] = n[i,t-1,g] + r[g]*n[i,t-1,g]*(1-n[i,t-1,g]/K)
    }
  }
}

stats_1.95 <- apply(n[,,1], 2, quantile,c(0.025,0.5,0.975))
stats_2.05 <- apply(n[,,2], 2, quantile,c(0.025,0.5,0.975))
stats_2.8 <- apply(n[,,3], 2, quantile,c(0.025,0.5,0.975))

plot(time,stats_1.95[2,],
     type = "l",
     xlab = "Time",
     ylab = "Population",
     main = "Initial Condition Uncertainty, r = 1.95")
ciEnvelope(time,stats_1.95[1,],stats_1.95[3,], col = "lightblue")
lines(time,stats_1.95[2,])

plot(time,stats_2.05[2,],
     type = "l",
     xlab = "Time",
     ylab = "Population",
     main = "Initial Condition Uncertainty, r = 2.05")
ciEnvelope(time,stats_2.05[1,],stats_2.05[3,], col = "lightblue")
lines(time,stats_2.05[2,])

plot(time,stats_2.8[2,],
     type = "l",
     xlab = "Time",
     ylab = "Population",
     main = "Initial Condition Uncertainty, r = 2.8")
ciEnvelope(time,stats_2.8[1,],stats_2.8[3,], col = "lightblue")
lines(time,stats_2.8[2,])

```

