---
title: "Foster_Chapter_11"
author: "John Foster"
date: "March 31, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r}
load("data/Ch11_UA.RData")
library(ecoforecastR)

logisticRE <- "
model{

## priors
r_global ~ dnorm(0,0.1)     ## across-site mean growth rate
K_global ~ dlnorm(6,0.01)   ## across-site mean carrying capacity
beta ~ dnorm(0,0.000001)    ## slope of K response to precip
tau_site ~ dgamma(0.1,0.1)  ## site random effect precision
R ~ dgamma(0.01,0.00000001) ## Observation error precision
Q ~ dgamma(0.01,0.00000001) ## Process errror precision 

## random effects and initial conditions, s = site
for(s in 1:NS){
  alpha_site[s] ~ dnorm(0,tau_site)  ## random site effect on K
  lN[s,1] ~ dnorm(6,0.001)           ## prior on IC, log scale
  N[s,1] <- exp(lN[s,1])             ## IC, linear scale
}

## process model, t = time, s = site
for(t in 2:NT){
  for(s in 1:NS){

  ## K is a linear model with a site random effect and fixed effect on log(precip)
  K[s,t]  <- max(1,K_global+alpha_site[s]+beta*log(precip[t]/800))  

  ## standard logistic growth process model, logged     
  mu[s,t] <- log(max(1,N[s,t-1] + r_global*N[s,t-1]*(1-N[s,t-1]/K[s,t])))

  ## process error
  lN[s,t] ~ dnorm(mu[s,t],Q)
  N[s,t] <- exp(lN[s,t])
  }
}
## observation model
for(t in 1:NT){
  for(s in 1:NS){
    No[s,t] ~ dlnorm(lN[s,t],R)
  }
}
}"

### jags output loaded in data ###
## parameters
# plot(out$params)
# summary(out$params)

### function to plot runs
plot.run <- function(){
  sel = seq(s,ncol(ci),by=NS)
  plot(time,time,type='n',ylim=ylim)
  ecoforecastR::ciEnvelope(time1,ci[1,sel],ci[3,sel],col=col.alpha("lightBlue",0.6))
  lines(time1,ci[2,sel],col="blue")
  points(time1,No[s,])
}

### plot settings
s <- 6             ## Focal site for forward simulation
Nmc = 1000         ## set number of Monte Carlo draws
ylim = c(100,700)  ## set Y range on plot
N.cols <- c("black","red","green","blue","orange") ## set colors
trans <- 0.8       ## set transparancy
time = 1:(NT*2)    ## total time
time1 = 1:NT       ## calibration period
time2 = time1+NT   ## forecast period

ci <- apply(as.matrix(out$predict),2,quantile,c(0.025,0.5,0.975))
plot.run()


##` @param IC    Initial Conditions
##` @param r     Intrinsic growth rate
##` @param Kg    Across-site ('global') mean carrying capacity
##` @param alpha Site random effect
##` @param beta  Slope of precipitation effect on K
##` @param ppt   Precipitation forecast
##` @param Q     Process error (default = 0 for deterministic runs)
##` @param n     Size of Monte Carlo ensemble
forecastN <- function(IC,r,Kg,alpha,beta,ppt,Q=0,n=Nmc){
  N <- matrix(NA,n,NT)  ## storage
  Nprev <- IC           ## initialize
  for(t in 1:NT){
    K = pmax(1,Kg + alpha + beta*log(ppt[,t]/800))  ## calculate carrying capacity
    mu = log(pmax(1,Nprev + r*Nprev*(1-Nprev/K)))   ## calculate mean
    N[,t] <- rlnorm(n,mu,Q)                         ## predict next step
    Nprev <- N[,t]                                  ## update IC
  }
  return(N)
}

## calculate mean of all inputs
ppt.mean <- matrix(apply(ppt_ensemble,2,mean),1,NT) ## driver
## parameters
params <- as.matrix(out$params)
param.mean <- apply(params,2,mean)
## initial conditions
IC <- as.matrix(out$predict)

N.det <- forecastN(IC=mean(IC[,"N[6,30]"]),
                   r=param.mean["r_global"],
                   Kg=param.mean["K_global"],
                   alpha=param.mean["alpha_site[6]"],
                   beta=param.mean["beta"],
                   ppt=ppt.mean,
                   Q=0,  ## process error off
                   n=1)

## Plot run
plot.run()
lines(time2,N.det,col="purple",lwd=3)

## sample parameter rows from previous analysis
prow = sample.int(nrow(params),Nmc,replace=TRUE)

N.I <- forecastN(IC=IC[prow,"N[6,30]"],  ## sample IC
                   r=param.mean["r_global"],
                   Kg=param.mean["K_global"],
                   alpha=param.mean["alpha_site[6]"],
                   beta=param.mean["beta"],
                   ppt=ppt.mean,
                   Q=0,
                   n=Nmc)

## Plot run
plot.run()
N.I.ci = apply(N.I,2,quantile,c(0.025,0.5,0.975))
ecoforecastR::ciEnvelope(time2,N.I.ci[1,],N.I.ci[3,],col=col.alpha(N.cols[1],trans))
lines(time2,N.I.ci[2,],lwd=0.5)
```

**Question 1:** Describe the pattern to the initial condition uncertainty and compare that to your expectations (e.g. Table 1).

> IC uncertainty plays a small role in the overall uncertainty at the beginning of the forecast, then reduces to essentially nothing after about 5 time steps.

```{r}
N.IP <- forecastN(IC=IC[prow,"N[6,30]"],  ## sample IC
                   r=params[prow,"r_global"],  ## sample parameters
                   Kg=params[prow,"K_global"],
                   alpha=params[prow,"alpha_site[6]"],
                   beta=params[prow,"beta"],
                   ppt=ppt.mean,
                   Q=0,
                   n=Nmc)

## Plot run
plot.run()
N.IP.ci = apply(N.IP,2,quantile,c(0.025,0.5,0.975))
ecoforecastR::ciEnvelope(time2,N.IP.ci[1,],N.IP.ci[3,],col=col.alpha(N.cols[2],trans))
ecoforecastR::ciEnvelope(time2,N.I.ci[1,],N.I.ci[3,],col=col.alpha(N.cols[1],trans))
lines(time2,N.I.ci[2,],lwd=0.5)
```

**Question 2:** 

A) Describe the pattern and relative magnitude of the parameter uncertainty and compare this to your expectations. 

> Parameter uncertainty is larger than IC uncertainty, and stays faily constant and normal around the IC mean throughout the forecast. This makes sense, as these parameters don't vary with time. 

B) Using the existing model and Monte Carlo approach, describe how you might partition the contribution of each parameter (r, Kg, alpha, beta) to the forecast uncertainty.

> Right now we are taking a random sample from all the parameters while holding ppt at the mean and not including process error. To further partition the parmeters, we can look at a random sample of just one of them while holding the rest at their mean. So to see what the relative contribution of growth rate we could run the simulation like this:

```
forecastN(IC=mean(IC[,"N[6,30]"]),  
                   r=params[prow,"r_global"],  ## sample just r
                   Kg=param.mean["K_global"],
                   alpha=param.mean["alpha_site[6]"],
                   beta=param.mean["beta"],
                   ppt=ppt.mean,
                   Q=0,
                   n=Nmc)
```

```{r}
## sample driver rows
drow = sample.int(nrow(ppt_ensemble),Nmc,replace=TRUE)

N.IPD <- forecastN(IC=IC[prow,"N[6,30]"],  ## sample IC
                   r=params[prow,"r_global"],  ## sample parameters
                   Kg=params[prow,"K_global"],
                   alpha=params[prow,"alpha_site[6]"],
                   beta=params[prow,"beta"],
                   ppt=ppt_ensemble[drow,],   ## Sample drivers
                   Q=0,
                   n=Nmc)

## Plot run
plot.run()
N.IPD.ci = apply(N.IPD,2,quantile,c(0.025,0.5,0.975))
ecoforecastR::ciEnvelope(time2,N.IPD.ci[1,],N.IPD.ci[3,],col=col.alpha(N.cols[3],trans))
ecoforecastR::ciEnvelope(time2,N.IP.ci[1,],N.IP.ci[3,],col=col.alpha(N.cols[2],trans))
ecoforecastR::ciEnvelope(time2,N.I.ci[1,],N.I.ci[3,],col=col.alpha(N.cols[1],trans))
lines(time2,N.I.ci[2,],lwd=0.5)
```

**Question 3:** As before, describe the pattern of how driver uncertainty contributes to the forecast and compare that to our first principles expectations.

> Driver uncertainty increases over time. Which is to be expected as the uncertainty in the ensamble forecast for the driver also increases over time.

```{r}
## process error samples
Qmc <- 1/sqrt(params[prow,"Q"])  ## convert from precision to standard deviation

N.IPDE <- forecastN(IC=IC[prow,"N[6,30]"],  ## sample IC
                   r=params[prow,"r_global"],  ## sample parameters
                   Kg=params[prow,"K_global"],
                   alpha=params[prow,"alpha_site[6]"],
                   beta=params[prow,"beta"],
                   ppt=ppt_ensemble[drow,],   ## Sample drivers
                   Q=Qmc,
                   n=Nmc)

## Plot run
plot.run()
N.IPDE.ci = apply(N.IPDE,2,quantile,c(0.025,0.5,0.975))
ecoforecastR::ciEnvelope(time2,N.IPDE.ci[1,],N.IPDE.ci[3,],col=col.alpha(N.cols[4],trans))
ecoforecastR::ciEnvelope(time2,N.IPD.ci[1,],N.IPD.ci[3,],col=col.alpha(N.cols[3],trans))
ecoforecastR::ciEnvelope(time2,N.IP.ci[1,],N.IP.ci[3,],col=col.alpha(N.cols[2],trans))
ecoforecastR::ciEnvelope(time2,N.I.ci[1,],N.I.ci[3,],col=col.alpha(N.cols[1],trans))
lines(time2,N.I.ci[2,],lwd=0.5)
```

**Question 4:** Why might the relative contribution of the process error decline with time? Does this match expectations?

> Because process error remains constant while driver error increases over time. Which is to be expected. Process error reflects our mis-representation of the process at hand, which will remain constant. While driver error increases because the uncertainty in driver forecast also increases.  


```{r}
## Random effect samples
tau.mc <- 1/sqrt(params[prow,"tau_site"]) ## converst from precision to std deviation
aNew.mc <- rnorm(Nmc,0,tau.mc)

N.IPDEA <- forecastN(IC=IC[prow,"N[6,30]"],  ## sample IC
                   r=params[prow,"r_global"],  ## sample parameters
                   Kg=params[prow,"K_global"],
                   alpha=aNew.mc,              ## sample random effect
                   beta=params[prow,"beta"],
                   ppt=ppt_ensemble[drow,],   ## Sample drivers
                   Q=Qmc,
                   n=Nmc)

## Plot run
plot.run()
N.IPDEA.ci = apply(N.IPDEA,2,quantile,c(0.025,0.5,0.975))
ecoforecastR::ciEnvelope(time2,N.IPDEA.ci[1,],N.IPDEA.ci[3,],col=col.alpha(N.cols[5],trans))
ecoforecastR::ciEnvelope(time2,N.IPDE.ci[1,],N.IPDE.ci[3,],col=col.alpha(N.cols[4],trans))
ecoforecastR::ciEnvelope(time2,N.IPD.ci[1,],N.IPD.ci[3,],col=col.alpha(N.cols[3],trans))
ecoforecastR::ciEnvelope(time2,N.IP.ci[1,],N.IP.ci[3,],col=col.alpha(N.cols[2],trans))
ecoforecastR::ciEnvelope(time2,N.I.ci[1,],N.I.ci[3,],col=col.alpha(N.cols[1],trans))
lines(time2,N.I.ci[2,],lwd=0.5)

### calculation of variances
varI     <- apply(N.I,2,var)
varIP    <- apply(N.IP,2,var)
varIPD   <- apply(N.IPD,2,var)
varIPDE  <- apply(N.IPDE,2,var)
varIPDEA <- apply(N.IPDEA,2,var)
varMat   <- rbind(varI,varIP,varIPD,varIPDE,varIPDEA)

## out-of-sample stacked area plot
V.pred.rel <- apply(varMat,2,function(x) {x/max(x)})
plot(time2,V.pred.rel[1,],ylim=c(0,1),type='n',main="Relative Variance: Out-of-Sample",ylab="Proportion of Variance",xlab="time")
ciEnvelope(time2,rep(0,ncol(V.pred.rel)),V.pred.rel[1,],col=N.cols[1])
ciEnvelope(time2,V.pred.rel[1,],V.pred.rel[2,],col=N.cols[2])
ciEnvelope(time2,V.pred.rel[2,],V.pred.rel[3,],col=N.cols[3])
ciEnvelope(time2,V.pred.rel[3,],V.pred.rel[4,],col=N.cols[4])
ciEnvelope(time2,V.pred.rel[4,],V.pred.rel[5,],col=N.cols[5])
legend("topleft",legend=c("RandomEffect","Process","Driver","Parameter","InitCond"),col=rev(N.cols),lty=1,lwd=5)

## in-sample stacked area plot
V.pred.rel.in <- apply(varMat[-5,],2,function(x) {x/max(x)})
plot(time2,V.pred.rel.in[1,],ylim=c(0,1),type='n',main="Relative Variance: In-Sample",ylab="Proportion of Variance",xlab="time")
ciEnvelope(time2,rep(0,ncol(V.pred.rel.in)),V.pred.rel.in[1,],col=N.cols[1])
ciEnvelope(time2,V.pred.rel.in[1,],V.pred.rel.in[2,],col=N.cols[2])
ciEnvelope(time2,V.pred.rel.in[2,],V.pred.rel.in[3,],col=N.cols[3])
ciEnvelope(time2,V.pred.rel.in[3,],V.pred.rel.in[4,],col=N.cols[4])
legend("topleft",legend=c("Process","Driver","Parameter","InitCond"),col=rev(N.cols[-5]),lty=1,lwd=5)
```


**Question 5:**  Based on results, where would you focus your research efforts next if your aim was to improve out-of-sample predictions? What about in-sample predictions?

> For out-of-sample predictions we should focus on reducing Random Effect uncertainty, which means constrainting site-to-site variability. For in-sample, we could focus on process or driver uncertainty. For process, we could try changing the underlying structure of the model. For driver we could try to find a different product, or just try a different variable. 



# Ensemble uncertainty propagation


```{r}

### Deterministic prediction ### 

## calculate mean of all inputs
ppt.mean <- matrix(apply(ppt_ensemble,2,mean),1,NT) ## driver
## parameters
params <- as.matrix(out$params)
param.mean <- apply(params,2,mean)
## initial conditions
IC <- as.matrix(out$predict)

N.det <- forecastN(IC=mean(IC[,"N[6,30]"]),
                   r=param.mean["r_global"],
                   Kg=param.mean["K_global"],
                   alpha=param.mean["alpha_site[6]"],
                   beta=param.mean["beta"],
                   ppt=ppt.mean,
                   Q=0,  ## process error off
                   n=1)

## Plot run
plot.run()
lines(time2,N.det,col="purple",lwd=3)

prow = sample.int(nrow(params),Nmc,replace=TRUE)
# IC uncertainty
N.I <- forecastN(IC=IC[prow,"N[6,30]"],  ## sample IC
                 r=param.mean["r_global"],
                 Kg=param.mean["K_global"],
                 alpha=param.mean["alpha_site[6]"],
                 beta=param.mean["beta"],
                 ppt=ppt.mean,
                 Q=0,
                 n=Nmc)


# function to calculate pred interval from mean and standard deviation
ens.pred <- function(dat){
              mean <- apply(dat, 2, mean)
              sd <- apply(dat, 2, sd)
              pred <- data.frame(up = mean + 1.96*sd,
                                 low = mean - 1.96*sd)
              return(pred)
}

pi.ic <- ens.pred(N.I)

plot.run()
ecoforecastR::ciEnvelope(time2,pi.ic[,1],pi.ic[,2],col=col.alpha(N.cols[1],trans))
lines(time2,pi.ic[,2],lwd=0.5)

# IC + param uncertainty

N.IP <- forecastN(IC=IC[prow,"N[6,30]"],  ## sample IC
                  r=params[prow,"r_global"],  ## sample parameters
                  Kg=params[prow,"K_global"],
                  alpha=params[prow,"alpha_site[6]"],
                  beta=params[prow,"beta"],
                  ppt=ppt.mean,
                  Q=0,
                  n=Nmc)

pi.IP <- ens.pred(N.IP)

plot.run()
ecoforecastR::ciEnvelope(time2,pi.IP[,1],pi.IP[,2],col=col.alpha(N.cols[2],trans))
ecoforecastR::ciEnvelope(time2,pi.ic[,1],pi.ic[,2],col=col.alpha(N.cols[1],trans))
lines(time2,pi.ic[,2],lwd=0.5)

# IC + param + driver uncertainty

## sample driver rows
drow = sample.int(nrow(ppt_ensemble),Nmc,replace=TRUE)

N.IPD <- forecastN(IC=IC[prow,"N[6,30]"],  ## sample IC
                   r=params[prow,"r_global"],  ## sample parameters
                   Kg=params[prow,"K_global"],
                   alpha=params[prow,"alpha_site[6]"],
                   beta=params[prow,"beta"],
                   ppt=ppt_ensemble[drow,],   ## Sample drivers
                   Q=0,
                   n=Nmc)

pi.IPD <- ens.pred(N.IPD)

plot.run()
ecoforecastR::ciEnvelope(time2,pi.IPD[,1],pi.IPD[,2],col=col.alpha(N.cols[3],trans))
ecoforecastR::ciEnvelope(time2,pi.IP[,1],pi.IP[,2],col=col.alpha(N.cols[2],trans))
ecoforecastR::ciEnvelope(time2,pi.ic[,1],pi.ic[,2],col=col.alpha(N.cols[1],trans))
lines(time2,pi.ic[,2],lwd=0.5)


# IC + param + driver + process uncertainty

## process error samples
Qmc <- 1/sqrt(params[prow,"Q"])  ## convert from precision to standard deviation

N.IPDE <- forecastN(IC=IC[prow,"N[6,30]"],  ## sample IC
                    r=params[prow,"r_global"],  ## sample parameters
                    Kg=params[prow,"K_global"],
                    alpha=params[prow,"alpha_site[6]"],
                    beta=params[prow,"beta"],
                    ppt=ppt_ensemble[drow,],   ## Sample drivers
                    Q=Qmc,
                    n=Nmc)

pi.IPDE <- ens.pred(N.IPDE)

plot.run()
ecoforecastR::ciEnvelope(time2,pi.IPDE[,1],pi.IPDE[,2],col=col.alpha(N.cols[4],trans))
ecoforecastR::ciEnvelope(time2,pi.IPD[,1],pi.IPD[,2],col=col.alpha(N.cols[3],trans))
ecoforecastR::ciEnvelope(time2,pi.IP[,1],pi.IP[,2],col=col.alpha(N.cols[2],trans))
ecoforecastR::ciEnvelope(time2,pi.ic[,1],pi.ic[,2],col=col.alpha(N.cols[1],trans))
lines(time2,pi.ic[,2],lwd=0.5)


# IC + param + driver + process + random effect uncertainty

tau.mc <- 1/sqrt(params[prow,"tau_site"]) ## converst from precision to std deviation
aNew.mc <- rnorm(Nmc,0,tau.mc)

N.IPDEA <- forecastN(IC=IC[prow,"N[6,30]"],  ## sample IC
                     r=params[prow,"r_global"],  ## sample parameters
                     Kg=params[prow,"K_global"],
                     alpha=aNew.mc,              ## sample random effect
                     beta=params[prow,"beta"],
                     ppt=ppt_ensemble[drow,],   ## Sample drivers
                     Q=Qmc,
                     n=Nmc)

pi.IPDEA <- ens.pred(N.IPDEA)

plot.run()
ecoforecastR::ciEnvelope(time2,pi.IPDEA[,1],pi.IPDEA[,2],col=col.alpha(N.cols[5],trans))
ecoforecastR::ciEnvelope(time2,pi.IPDE[,1],pi.IPDE[,2],col=col.alpha(N.cols[4],trans))
ecoforecastR::ciEnvelope(time2,pi.IPD[,1],pi.IPD[,2],col=col.alpha(N.cols[3],trans))
ecoforecastR::ciEnvelope(time2,pi.IP[,1],pi.IP[,2],col=col.alpha(N.cols[2],trans))
ecoforecastR::ciEnvelope(time2,pi.ic[,1],pi.ic[,2],col=col.alpha(N.cols[1],trans))
lines(time2,pi.ic[,2],lwd=0.5)

##### Uncertainty Analysis

### calculation of variances
varI     <- apply(N.I,2,var)
varIP    <- apply(N.IP,2,var)
varIPD   <- apply(N.IPD,2,var)
varIPDE  <- apply(N.IPDE,2,var)
varIPDEA <- apply(N.IPDEA,2,var)
varMat   <- rbind(varI,varIP,varIPD,varIPDE,varIPDEA)

## out-of-sample stacked area plot
V.pred.rel <- apply(varMat,2,function(x) {x/max(x)})
plot(time2,V.pred.rel[1,],ylim=c(0,1),type='n',main="Relative Variance: Out-of-Sample",ylab="Proportion of Variance",xlab="time")
ciEnvelope(time2,rep(0,ncol(V.pred.rel)),V.pred.rel[1,],col=N.cols[1])
ciEnvelope(time2,V.pred.rel[1,],V.pred.rel[2,],col=N.cols[2])
ciEnvelope(time2,V.pred.rel[2,],V.pred.rel[3,],col=N.cols[3])
ciEnvelope(time2,V.pred.rel[3,],V.pred.rel[4,],col=N.cols[4])
ciEnvelope(time2,V.pred.rel[4,],V.pred.rel[5,],col=N.cols[5])
legend("topleft",legend=c("RandomEffect","Process","Driver","Parameter","InitCond"),col=rev(N.cols),lty=1,lwd=5)


## in-sample stacked area plot
V.pred.rel.in <- apply(varMat[-5,],2,function(x) {x/max(x)})
plot(time2,V.pred.rel.in[1,],ylim=c(0,1),type='n',main="Relative Variance: In-Sample",ylab="Proportion of Variance",xlab="time")
ciEnvelope(time2,rep(0,ncol(V.pred.rel.in)),V.pred.rel.in[1,],col=N.cols[1])
ciEnvelope(time2,V.pred.rel.in[1,],V.pred.rel.in[2,],col=N.cols[2])
ciEnvelope(time2,V.pred.rel.in[2,],V.pred.rel.in[3,],col=N.cols[3])
ciEnvelope(time2,V.pred.rel.in[3,],V.pred.rel.in[4,],col=N.cols[4])
legend("topleft",legend=c("Process","Driver","Parameter","InitCond"),col=rev(N.cols[-5]),lty=1,lwd=5)



```


> The first thing that jumps out is that the uncertainty (sans random effect) is always centered around the median prediction, which is not the case in the orginal forecast. The random effect looks about the same, and the relative variance in- and out-of-sample look about the same. 

> In the ensamble below, process error representws a larger proportion of te overall uncertianty, while driver error is less. We also see more random effect uncertainty in this simulation. 

# Ensamble with Nmc/40

```{r}
Nmc <- Nmc/40

### Deterministic prediction ### 

## calculate mean of all inputs
ppt.mean <- matrix(apply(ppt_ensemble,2,mean),1,NT) ## driver
## parameters
params <- as.matrix(out$params)
param.mean <- apply(params,2,mean)
## initial conditions
IC <- as.matrix(out$predict)

N.det <- forecastN(IC=mean(IC[,"N[6,30]"]),
                   r=param.mean["r_global"],
                   Kg=param.mean["K_global"],
                   alpha=param.mean["alpha_site[6]"],
                   beta=param.mean["beta"],
                   ppt=ppt.mean,
                   Q=0,  ## process error off
                   n=1)

## Plot run
plot.run()
lines(time2,N.det,col="purple",lwd=3)

prow = sample.int(nrow(params),Nmc,replace=TRUE)
# IC uncertainty
N.I <- forecastN(IC=IC[prow,"N[6,30]"],  ## sample IC
                 r=param.mean["r_global"],
                 Kg=param.mean["K_global"],
                 alpha=param.mean["alpha_site[6]"],
                 beta=param.mean["beta"],
                 ppt=ppt.mean,
                 Q=0,
                 n=Nmc)


# function to calculate pred interval from mean and standard deviation
ens.pred <- function(dat){
              mean <- apply(dat, 2, mean)
              sd <- apply(dat, 2, sd)
              pred <- data.frame(up = mean + 1.96*sd,
                                 low = mean - 1.96*sd)
              return(pred)
}

pi.ic <- ens.pred(N.I)

plot.run()
ecoforecastR::ciEnvelope(time2,pi.ic[,1],pi.ic[,2],col=col.alpha(N.cols[1],trans))
lines(time2,pi.ic[,2],lwd=0.5)

# IC + param uncertainty

N.IP <- forecastN(IC=IC[prow,"N[6,30]"],  ## sample IC
                  r=params[prow,"r_global"],  ## sample parameters
                  Kg=params[prow,"K_global"],
                  alpha=params[prow,"alpha_site[6]"],
                  beta=params[prow,"beta"],
                  ppt=ppt.mean,
                  Q=0,
                  n=Nmc)

pi.IP <- ens.pred(N.IP)

plot.run()
ecoforecastR::ciEnvelope(time2,pi.IP[,1],pi.IP[,2],col=col.alpha(N.cols[2],trans))
ecoforecastR::ciEnvelope(time2,pi.ic[,1],pi.ic[,2],col=col.alpha(N.cols[1],trans))
lines(time2,pi.ic[,2],lwd=0.5)

# IC + param + driver uncertainty

## sample driver rows
drow = sample.int(nrow(ppt_ensemble),Nmc,replace=TRUE)

N.IPD <- forecastN(IC=IC[prow,"N[6,30]"],  ## sample IC
                   r=params[prow,"r_global"],  ## sample parameters
                   Kg=params[prow,"K_global"],
                   alpha=params[prow,"alpha_site[6]"],
                   beta=params[prow,"beta"],
                   ppt=ppt_ensemble[drow,],   ## Sample drivers
                   Q=0,
                   n=Nmc)

pi.IPD <- ens.pred(N.IPD)

plot.run()
ecoforecastR::ciEnvelope(time2,pi.IPD[,1],pi.IPD[,2],col=col.alpha(N.cols[3],trans))
ecoforecastR::ciEnvelope(time2,pi.IP[,1],pi.IP[,2],col=col.alpha(N.cols[2],trans))
ecoforecastR::ciEnvelope(time2,pi.ic[,1],pi.ic[,2],col=col.alpha(N.cols[1],trans))
lines(time2,pi.ic[,2],lwd=0.5)


# IC + param + driver + process uncertainty

## process error samples
Qmc <- 1/sqrt(params[prow,"Q"])  ## convert from precision to standard deviation

N.IPDE <- forecastN(IC=IC[prow,"N[6,30]"],  ## sample IC
                    r=params[prow,"r_global"],  ## sample parameters
                    Kg=params[prow,"K_global"],
                    alpha=params[prow,"alpha_site[6]"],
                    beta=params[prow,"beta"],
                    ppt=ppt_ensemble[drow,],   ## Sample drivers
                    Q=Qmc,
                    n=Nmc)

pi.IPDE <- ens.pred(N.IPDE)

plot.run()
ecoforecastR::ciEnvelope(time2,pi.IPDE[,1],pi.IPDE[,2],col=col.alpha(N.cols[4],trans))
ecoforecastR::ciEnvelope(time2,pi.IPD[,1],pi.IPD[,2],col=col.alpha(N.cols[3],trans))
ecoforecastR::ciEnvelope(time2,pi.IP[,1],pi.IP[,2],col=col.alpha(N.cols[2],trans))
ecoforecastR::ciEnvelope(time2,pi.ic[,1],pi.ic[,2],col=col.alpha(N.cols[1],trans))
lines(time2,pi.ic[,2],lwd=0.5)


# IC + param + driver + process + random effect uncertainty

tau.mc <- 1/sqrt(params[prow,"tau_site"]) ## converst from precision to std deviation
aNew.mc <- rnorm(Nmc,0,tau.mc)

N.IPDEA <- forecastN(IC=IC[prow,"N[6,30]"],  ## sample IC
                     r=params[prow,"r_global"],  ## sample parameters
                     Kg=params[prow,"K_global"],
                     alpha=aNew.mc,              ## sample random effect
                     beta=params[prow,"beta"],
                     ppt=ppt_ensemble[drow,],   ## Sample drivers
                     Q=Qmc,
                     n=Nmc)

pi.IPDEA <- ens.pred(N.IPDEA)

plot.run()
ecoforecastR::ciEnvelope(time2,pi.IPDEA[,1],pi.IPDEA[,2],col=col.alpha(N.cols[5],trans))
ecoforecastR::ciEnvelope(time2,pi.IPDE[,1],pi.IPDE[,2],col=col.alpha(N.cols[4],trans))
ecoforecastR::ciEnvelope(time2,pi.IPD[,1],pi.IPD[,2],col=col.alpha(N.cols[3],trans))
ecoforecastR::ciEnvelope(time2,pi.IP[,1],pi.IP[,2],col=col.alpha(N.cols[2],trans))
ecoforecastR::ciEnvelope(time2,pi.ic[,1],pi.ic[,2],col=col.alpha(N.cols[1],trans))
lines(time2,pi.ic[,2],lwd=0.5)

##### Uncertainty Analysis

### calculation of variances
varI     <- apply(N.I,2,var)
varIP    <- apply(N.IP,2,var)
varIPD   <- apply(N.IPD,2,var)
varIPDE  <- apply(N.IPDE,2,var)
varIPDEA <- apply(N.IPDEA,2,var)
varMat   <- rbind(varI,varIP,varIPD,varIPDE,varIPDEA)

## out-of-sample stacked area plot
V.pred.rel <- apply(varMat,2,function(x) {x/max(x)})
plot(time2,V.pred.rel[1,],ylim=c(0,1),type='n',main="Relative Variance: Out-of-Sample",ylab="Proportion of Variance",xlab="time")
ciEnvelope(time2,rep(0,ncol(V.pred.rel)),V.pred.rel[1,],col=N.cols[1])
ciEnvelope(time2,V.pred.rel[1,],V.pred.rel[2,],col=N.cols[2])
ciEnvelope(time2,V.pred.rel[2,],V.pred.rel[3,],col=N.cols[3])
ciEnvelope(time2,V.pred.rel[3,],V.pred.rel[4,],col=N.cols[4])
ciEnvelope(time2,V.pred.rel[4,],V.pred.rel[5,],col=N.cols[5])
legend("topleft",legend=c("RandomEffect","Process","Driver","Parameter","InitCond"),col=rev(N.cols),lty=1,lwd=5)


## in-sample stacked area plot
V.pred.rel.in <- apply(varMat[-5,],2,function(x) {x/max(x)})
plot(time2,V.pred.rel.in[1,],ylim=c(0,1),type='n',main="Relative Variance: In-Sample",ylab="Proportion of Variance",xlab="time")
ciEnvelope(time2,rep(0,ncol(V.pred.rel.in)),V.pred.rel.in[1,],col=N.cols[1])
ciEnvelope(time2,V.pred.rel.in[1,],V.pred.rel.in[2,],col=N.cols[2])
ciEnvelope(time2,V.pred.rel.in[2,],V.pred.rel.in[3,],col=N.cols[3])
ciEnvelope(time2,V.pred.rel.in[3,],V.pred.rel.in[4,],col=N.cols[4])
legend("topleft",legend=c("Process","Driver","Parameter","InitCond"),col=rev(N.cols[-5]),lty=1,lwd=5)



```